### 2025-01-16 sections/1-introduction.tex
more edits after another reading

Researchers are collecting data at an unprecedented rate.
In many fields, datasets grow exponentially, and this increase in the rate of data collection outpaces improvements in computing performance as predicted by Moore's Law~\cite{kahn2011future}.
This indicates that the performance of computing systems will not keep pace with the growth of data.
Often dubbed ``the Big Data explosion,'' this phenomenon has created a need for better algorithms to analyze large datasets.

>>>>

Researchers are collecting data at an unprecedented rate.
In many fields, datasets are growing exponentially, and this increase in the rate of data collection outpaces improvements in computing performance as predicted by Moore's Law~\cite{kahn2011future}.
This indicates that the performance of computing systems will not keep pace with the growth of data.
Often dubbed ``the Big Data explosion,'' this phenomenon has created a need for better algorithms to analyze large datasets.

### 2025-01-16 sections/3-datasets.tex
more edits after another reading

We benchmarked CAKES's algorithms against the state-of-the-art similarity search algorithms HNSW, ANNOY and FAISS-IVF, and two implementations of linear search.
In particular, we use FAISS-Flat, a Python implementation of linear search, to provide ground-truth values for recall of HNSW, ANNOY, and FAISS-IVF.
We use our own Rust implementation of linear search to provide ground-truth values for calculating recall of CAKES's algorithms.
We plot the throughput of these two implementations of linear search as a reference point for all other algorithms in Figure~\ref{fig:results:scaling-plots}.

>>>>

We benchmarked the three CAKES algorithms against a na\"ive linear search implementation in Rust.
We also benchmarked against three state-of-the-art similarity search algorithms: HNSW, ANNOY, and FAISS-IVF in Python.
We verified that our implementation of linear search produces the same neighbors as provided by the ANN-Benchmarks suite for Fashion-MNIST, Glove-25 and Sift datasets.
We then used this linear search implementation to find and store the ground-truth for the augmented versions of the datasets.
We used this ground-truth to calculate recall for CAKES's algorithms in Rust and for HNSW, ANNOY, and FAISS-IVF in Python.
We plot the results of these benchmarks in Figure~\ref{fig:results:scaling-plots}.

### 2025-01-16 sections/4-results.tex
more edits after another reading

The Silva-18S dataset consists of genomic sequences whose unaligned lengths are at-most 3,712 but were provided in an MSA of width 50,000.
As such the embedding dimension is somewhere between 3,712 and 50,000.
We use the Levenshtein edit distance (a metric) to measure distances between sequences.
This dataset, as shown in Figure~\ref{fig:results:silva-lfd}, exhibits consistently low LFD.
In particular, LFD is less than about 3 for all depths, hovering near 1 for clusters at depth 40 and deeper.

The Radio-ML dataset consists of measurements of radio-frequency signals using 1,024 dimensional complex-valued vectors.
We use the Dynamic Time Warping distance metric on this dataset.
This dataset is synthetic~\cite{oshea2018radioml} but uses a far more elaborate generation process than our Random dataset.
The LFD values show three distinct peaks around an LFD of 12 at or near depths of 8, 25 and 50.
Each peak is followed by a linear decrease until encountering a sharp spike for the next peak.
Within each of the tree portions, this dataset has a character very similar to that of the Random dataset.
This suggests that the dataset obeys the manifold hypothesis at some scales, but that it is not ``scale free,'' as the LFD varies significantly by depth;
This is likely the result of a piecewise uniform sampling strategy used to generate the different modulation modes present in the dataset.

@@

These four comparisons (referred to as Ball, BalancedBall, PermutedBall, and, PermutedBalancedBall, where depth-first-reordering is referred to as Permuted for brevity) for all three search algorithms on the Fashion-MNIST dataset are shown in Figure~\ref{fig:results:distance-counts}.
Notably, performance is not only strictly worse for the balanced approaches, but the asymptotic behavior is significantly worse;
the instances (such as Depth-First Sieve) that seemed to exhibit constant-time behavior in Figure~\ref{fig:results:scaling-plots}, which is effectively PermutedBall, no longer do under a balanced clustering.
We note that on this dataset, depth-first reordering (the Permuted plotlines) seems to have little effect; depth-first reordering is intended to improve space complexity only.

>>>>

The Silva-18S dataset consists of genomic sequences whose unaligned lengths are at-most 3,712 but were provided in an MSA of width 50,000.
As such the embedding dimension is somewhere between 3,712 and 50,000.
We use the Levenshtein edit distance (a metric) to measure distances between sequences.
This dataset, as shown in Figure~\ref{fig:results:silva-lfd}, exhibits consistently low LFD.
In particular, LFD is less than 3 for all depths, hovering near 1 for clusters at depth 40 and deeper.

The Radio-ML dataset consists of measurements of radio-frequency signals using 1,024 dimensional complex-valued vectors.
We use the Dynamic Time Warping distance metric on this dataset.
This dataset is synthetic~\cite{oshea2018radioml} but uses a far more elaborate generation process than our Random dataset.
The LFD values show three distinct peaks around an LFD of 12 at or near depths of 8, 25 and 50.
Each peak is followed by a linear decrease until encountering a sharp spike for the next peak.
Within each of the tree portions, this dataset has a character very similar to that of the Random dataset.
This suggests that the dataset obeys the manifold hypothesis at some scales, but that it is not ``scale free,'' as the LFD varies significantly by depth.
This is likely the result of a piecewise uniform sampling strategy used to generate the different modulation modes present in the dataset.

@@

These four comparisons (referred to as Ball, BalancedBall, PermutedBall, and, PermutedBalancedBall, where depth-first-reordering is referred to as Permuted for brevity) for all three search algorithms on the Fashion-MNIST dataset are shown in Figure~\ref{fig:results:distance-counts}.
Notably, performance is not only strictly worse for the balanced approaches, but the asymptotic behavior is significantly worse;
the algorithms (such as Depth-First Sieve) that seemed to exhibit constant-time behavior in Figure~\ref{fig:results:scaling-plots}, which shows results using PermutedBall, no longer do so under a balanced clustering.
We note that on this dataset, depth-first reordering (the Permuted variants) seems to have little effect on throughput; depth-first reordering is primarily intended to improve space complexity.

### 2025-01-16 sections/5-discussion.tex
more edits after another reading

When written with the same notation as used in Section~\ref{sec:methods:knn-search:complexity-of-sieve-methods}, we see that the time complexity of Repeated $\rho$-NN is $\mathcal{O}(\tfrac{T}{\lceil d \rceil} + L)$, where $T$ is the time complexity of tree-search and $L$ is the time complexity of leaf-search.
Even though Repeated $\rho$-NN has the least time cost (compare to $\mathcal{O}(T\log{(T+L)} + L\log{(T+L)})$ for Breadth-First Sieve and $\mathcal{O}(T\log{T} + L\log{k})$ for Depth-First Sieve) of the three CAKES algorithms, it is not always the fastest algorithm empirically.
We believe that some of this discrepancy can be explained by the fact that if Repeated $\rho$-NN significantly ``overshoots'' the correct radius for $k$ hits ($\rho_k$) during tree-search, leaf-search will require looking at more points, rendering the true scaling factor higher than that in Equation~\ref{eq:methods:repeated-rnn-complexity}.
This overshot can occur when the LFDs of clusters near the query are not concentrated around their expectation.
For example, if most clusters near the query have very low LFD except for one anomaly with very high LFD,
the harmonic mean LFD $\mu$ can still be low, so the factor of radial increase in (Equation)~\ref{eq:methods:repeated-rnn-factor} may be much larger than necessary for guaranteeing $k$ hits.
This suggests that rather than using the reciprocal of the harmonic mean LFD in~\ref{eq:methods:repeated-rnn-factor}, we may achieve better results with a mean that is more sensitive to high outliers, such as the geometric mean.
We leave it as an avenue for future work to characterize when Repeated $\rho$-NN significantly overestimates the correct radius $\rho_k$ and to improve upon the factor of radial increase in Equation~\ref{eq:methods:repeated-rnn-factor} so that this occurs less frequently and with less severity.

While the fastest CAKES algorithm varies by dataset, we note that on all three ANN-benchmark datasets (Fashion-Mnist, Glove-25, and Sift), Depth-First Sieve and Breadth-First Sieve show nearly constant throughput as the cardinality increases.
This observation warrants further investigation, but it is especially promising given that both algorithms consistently outperform linear search for high cardinalities.
Finally, we emphasize that the variation in performance of our algorithms across different datasets and cardinalities supports our use of an auto-tuning function (as discussed in Section~\ref{sec:methods:auto-tuning}) to select the best algorithm for a given dataset.
Future work is needed to better characterize those datasets for which each algorithm performs best, but our results suggest that a more sophisticated auto-tuning function could be developed to select the fastest algorithm based on the properties of the dataset.

@@

We note that despite the genericity of CAKES, effectively applying it to real-world datasets may typically require domain-specific knowledge.
As an example, consider the domain of protein sequence search; protein sequences are of highly variable length (30--1000 amino acids are typical, with some outliers) and typically comprise a 20-letter alphabet.
There have been numerous protein sequence search algorithms~\cite{kim2021entrance, daniels2013compressive, yu2015entropy, steinegger2018clustering}, and the one thing they have in common besides the application domain is that their development required deep domain-specific knowledge.
While simple edit distance may be a poor choice for clustering when sequence lengths exhibit such variance, other choices of distance metrics and preprocessing could be used by CAKES, such as representing each sequence as a set of $k$-mers (short substrings of length $k$) and choosing Jaccard distance as the metric; this approach was chosen by~\cite{kim2021entrance}.

The questions raised by this study suggest several additional avenues for future work.
A comparison across more datasets is in order, as is further analysis with other distance functions that existing methods, such as FAISS, HNSW and ANNOY, do not support.
These include Wasserstein distance~\cite{vallender1974calculation} for probability distributions (particularly for high dimensional distributions) and Tanimoto distance~\cite{bajusz2015tanimoto} for comparing molecular structures by their maximal common subgraphs.
Incorporating these or other distance functions in CAKES requires only that a Rust implementation of the distance function be provided.

>>>>

When written with the same notation as used in Section~\ref{sec:methods:knn-search:complexity-of-sieve-methods}, we see that the time complexity of Repeated $\rho$-NN is $\mathcal{O}(\tfrac{T}{\lceil d \rceil} + L)$, where $T$ is the time complexity of tree-search and $L$ is the time complexity of leaf-search.
Even though Repeated $\rho$-NN has the least time cost (compared to $\mathcal{O}\left((T + L)\log{(T+L)}\right)$ for Breadth-First Sieve and $\mathcal{O}(T\log{T} + L\log{k})$ for Depth-First Sieve) of the three CAKES algorithms, it is not always the fastest algorithm empirically.
We believe that some of this discrepancy can be explained by the fact that Repeated $\rho$-NN can significantly ``overshoots'' the correct radius for $k$ hits ($\rho_k$) during tree-search.
This causes leaf-search to exhaustively search many more clusters than necessary, rendering the true scaling factor higher than that in Equation~\ref{eq:methods:repeated-rnn-complexity}.
This overshot can occur when the LFDs of clusters near the query are not concentrated around their expectation.
For example, if most clusters near the query have very low LFD except for one anomaly with very high LFD,
the harmonic mean LFD $\mu$ can still be low, so the factor of radial increase in (Equation)~\ref{eq:methods:repeated-rnn-factor} may be much larger than necessary for guaranteeing $k$ hits.
This suggests that rather than using the reciprocal of the harmonic mean LFD in~\ref{eq:methods:repeated-rnn-factor}, we may achieve better results with a mean that is more sensitive to high outliers, such as the geometric mean.
We leave it as an avenue for future work to characterize when Repeated $\rho$-NN significantly overestimates the correct radius $\rho_k$ and to improve upon the factor of radial increase in Equation~\ref{eq:methods:repeated-rnn-factor} so that this occurs less frequently and with less severity.

While the fastest CAKES algorithm can vary by dataset and cardinality, we note that on all three ANN-benchmark datasets (Fashion-MNIST, Glove-25, and Sift), Depth-First Sieve and Breadth-First Sieve show nearly constant throughput as the cardinality increases.
This observation warrants further investigation, but it is especially promising given that both algorithms consistently outperform linear search for high cardinalities.
Finally, we emphasize that the variation in performance of our algorithms across different datasets and cardinalities supports our use of an auto-tuning function (as discussed in Section~\ref{sec:methods:auto-tuning}) to select the best algorithm for a given dataset.
Future work is needed to better characterize those datasets for which each algorithm performs best, but our results suggest that a more sophisticated auto-tuning function could be developed to select the fastest algorithm based on the properties of the dataset.

@@

We note that despite the genericity of CAKES, effectively applying it to real-world datasets may require domain-specific knowledge.
As an example, consider the domain of protein sequence search; protein sequences are of highly variable length (30--1000 amino acids are typical, with some outliers) and typically comprise a 20-letter alphabet.
There have been numerous protein sequence search algorithms~\cite{kim2021entrance, daniels2013compressive, yu2015entropy, steinegger2018clustering}, and the one thing they have in common besides the application domain is that their development required deep domain-specific knowledge.
While simple edit distance may be a poor choice for clustering when sequence lengths exhibit such variance, other choices of distance metrics and preprocessing could be used by CAKES, such as representing each sequence as a set of $k$-mers (short substrings of length $k$) and choosing Jaccard distance on these sets as the metric; this approach was chosen by~\cite{kim2021entrance}.
If the distance function uses a global sequence alignment algorithm, such as Needleman-Wunsch~\cite{needleman1970general}, the dataset may also be preprocessed into a number of mutually exclusive subsets based on the lengths of the sequences, and a separate CAKES tree may be built for each subset.
At search time, the query sequence would be assigned to the appropriate tree based on its length, and the search would be performed on that tree, and on a small number of trees on adjacent lengths' bins.

The questions raised by this study suggest several additional avenues for future work.
A comparison across more datasets is in order, as is further analysis with other distance functions that existing methods, such as FAISS, HNSW and ANNOY, do not support.
These include Wasserstein distance~\cite{vallender1974calculation} for probability distributions (particularly for high dimensional distributions) and Tanimoto distance~\cite{bajusz2015tanimoto} for comparing molecular structures by their maximal common subgraphs.
Incorporating new distance functions in CAKES requires only that a Rust implementation of the distance function be provided.
